---
title: "Does Mileage depend on transmission type"
author: "Shubham Kalra for Auto Magazine"
email: "iam@shubhamkalra.com"
date: "Thursday, April 09, 2015"
output: html_document
---
=======================================

In the 1974 Motor Trend US magazine came out with datset which comprises fuel consumptions and 10 other aspects of automobile performance for 32 automobiles. Today, We look back to find parameters which determined the mileage of a car in that time. And quantify if cars with manual transmission or automatic transmission had a better mileage.  
  
The article is written keeping in mind audience with statistical know how.  One can replicate the study using code snippets provided alongside. To keep the analysis more organised and less overwheling i have divided into various sub-sections. Lets jump right into it. 


### Exploratory Data Analysis


The data comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

**mtcars is data frame with 32 observations on 11 variables**

Variable | Definition
------ | ----------------
mpg  |  Miles/(US) gallon  
cyl  |	Number of cylinders  
disp	|	Displacement (cu.in.)  
hp	|	Gross horsepower  
drat	|	Rear axle ratio  
wt	|	Weight (lb/1000)  
qsec	|	1/4 mile time  
vs	|	V/S  
am	|	Transmission (0 = automatic, 1 = manual)  
gear	|	Number of forward gears  
carb	|	Number of carburetors	


-------------------------------------------------------------------

We start with Importing Data into a new variable, say `cars`, and displaying first few observations of `cars` 

```{r}
data(mtcars)
#importing data in a variable
cars <- mtcars
head(cars)

```

One should also run `summary(cars)` to get more intresting stats about each variable. We see that a few variables like, Number of Cyclinders, Number of forward gears etc. are categorical variables. We convert those to something called as [factor variables](http://www.stat.berkeley.edu/~s133/factors.html) 


```{r}

# Changing classifying variables  from binary values to factor

cars$cyl <- factor(cars$cyl)
cars$vs <- factor(cars$vs)
cars$am <- factor(cars$am)
cars$gear <- factor(cars$gear)
cars$carb <- factor(cars$carb)
```



### Data Visualizations

Next, we visualize the data to understand the behaviour of our response variable. 


```{r, echo=FALSE}
par(mfrow = c(1, 2)) 
#visualize two graphs in one horizontal length
# Add a Normal Curve (Thanks to Peter Dalgaard)
x <- cars$mpg 
h<-hist(x, breaks=10, col="cornflowerblue", xlab="Miles Per Gallon", 
    main="Histogram with Normal Curve") 
xfit<-seq(min(x),max(x),length=40) 
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x)) 
yfit <- yfit*diff(h$mids[1:2])*length(x) 
lines(xfit, yfit, col="blue", lwd=2)

# Density Curve
plot(density(cars$mpg), xlab = "MPG", main ="Density Plot of MPG")
```

* Curve shows a near Normal Distribution


```{r, echo=FALSE, message=FALSE, warning=FALSE}
nums <- sapply(cars, is.numeric)

pairs(cars[,nums], main ="Finding Correlation among Variables", panel=function(x,y) {
  points(x, y)
  abline(lm(y ~ x), col="red")
})

```

* Above is a pairs plot, which is used for drawing a scatter plot matrix to visually study the relationships between two variables.
* We see a high correlation of mpg with many variables. 
* Also, we see many correlations among predictor variables. 


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plyr)
library(ggplot2)
# Rename the levels of transmission types
transmission <- revalue(cars$am, c('0'="automatic", '1'="manual"))
qplot(transmission, mpg, data=cars, geom=c("boxplot"), 
   fill=am, main="Mileage by Transmission Type",
   xlab="Transmission Type", ylab="Miles per Gallon")

```

from the box plot, we see.  

* There are no real outliers in the dataset. 
* Manual transmission is better than Automatic transmission, next we will check if they are statistically significant using a [two sample t-test](http://www.itl.nist.gov/div898/handbook/eda/section3/eda353.htm)  
  
  
Don't worry about the code for the above plots, you can fork the entire code on github [here](https://github.com/shubhamkalra27/mtcars-analysis/blob/master/mtcars.R) 
   
  
  
#### Two Sample t-test

Two Sample t-test is used to compare the means of a normally distributed interval dependent variable for two independent groups.

``` {r}
t.test(mpg~am, data = cars)
```

**Interpretation**   
From the above output, we reject the null hypothesis With a p-value of **0.001374**,  and say that there is a statistically significant difference in mean mileage between Manual transmission and automatic transmission.

We will now Quantify that difference using Regression techniques.


###Simple linear regression

Our first regression model is a model to explain the `mpg` variability with the `am` only.  


``` {r}
model1 <- lm( mpg~am ,data=mtcars); summary(model1)
```


**Interpretation of simple linear regression**  

* Coefficient value of am says that Automatic Transmission has **`r round(summary(model1)$coefficients[1],3)`** average MPG  
* Intercepts value means that Manual transmission cars add **`r round(summary(model1)$coefficients[2],3)`** MPG's more to the Automatic's mean value  
* Adj R square value says, that only **`r round(summary(model1)$adj.r.squared*100,2)`%** of deviation in model can be explained by the current model with one predictor.
  


> Before making any conclusions on the effect of transmission type on mileage, we look at the impact of other variables on the mileage



  
### Multiple Regression

Next, we would like to see dependence of other variables on Mileage. To find out which variables to use in our analysis, We run a Stepwise regression to identify useful subset of predictors. 

```{r, eval=FALSE }
nullModel <- lm (mpg ~ 1 , data =cars) # model with no predictor variables.

fullModel <- lm (mpg ~ . , data = cars) # model with all variables added as predictor variables. 

bestModel <- step(fullModel, nullModel, direction = "both" , k=  log(nrow(cars))) #both direction BIC step model

bestModel$call  #display formula
```

```{r, echo=FALSE }
nullModel <- lm (mpg ~ 1 , data =cars) # model with no predictor variables.

fullModel <- lm (mpg ~ . , data = cars) # model with all variables added as predictor variables. 

bestModel <- step(fullModel, nullModel, direction = "both", k=  log(nrow(cars))) #both direction step model

bestModel$call  #display formula
```


* The variables mentioned by bestModel are uncorrelated and most useful set of predictors
* The final model `bestModel` contains three predictors, wt(weight), qsec(1/4 mile time) and am(Transmission)  
 

Summarizing the new Model.

```{r}
summary(bestModel)
```
  
  
  
**Interpretation of Multiple Regression**  

  
* Adjusted R square value says, that **`r round(summary(bestModel)$adj.r.squared*100,2)`%** of the variance can be explained by this model.Which is a significant improvement over first model.  
  
* Coefficient for `am` says that, on average, manual transmission cars have **`r round(summary(bestModel)$coefficients[4],3)`**  MPGs more than automatic transmission cars. 

* The p-value for `am` is `0.046` which makes it a weakest predictor of mileage.

-------------------------------------------------------------------

 
### Conclusion

When seen with just one predictor `am` we get Manual transmission to yield `7.245`mpg higher than automatic transmission. Only **`r round(summary(model1)$adj.r.squared*100,2)`%** of variance could be explained in this case.

With addition of more predictors, difference in mileage between manual transmission and automatic transmission is just `2.936`mpg. The p-value significance of transmission is also very low. Weight and qsec have [confounded](http://en.wikipedia.org/wiki/Confounding/ "wiki article") the relationship between transmission and mpg, this model explains over **`r round(summary(bestModel)$adj.r.squared*100,2)`%** of the variance. 


You can access the code for the above analysis at my github repository. <https://github.com/shubhamkalra27/mtcars-analysis>

-----------------------------------------  
  
#### References. 

1. mtcars dataset <https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html>
2. Density Plot. <http://www.statmethods.net/graphs/density.html>   
3. Analysis on mtcars dataset to predict mpg <http://varianceexplained.org/RData/code/code_lesson3/>
4. BoxPlot using ggplot2 <http://www.statmethods.net/advgraphs/ggplot2.html>
5. More on factors <http://www.stat.berkeley.edu/~s133/factors.html>  
6. Confounding Variables <http://en.wikipedia.org/wiki/Confounding/>
