---
title: "Does Mileage depend on transmission type"
author: "Shubham Kalra"
email: "iam@shubhamkalra.com"
date: "Friday, April 10, 2015"
output: html_document
---
=======================================

In 1974, Motor Trend US magazine came out with data set which comprised of fuel consumptions and 10 other aspects of automobile performance for 32 automobiles. Today, We look back to find parameters which determined the mileage of a car at that point of time. And quantify, whether cars with manual transmission or cars with automatic transmission had a better mileage. 
 
This article is written keeping in mind audience with statistical know how. One can replicate the study using code snippets provided alongside. To keep the analysis more organized and less overwhelming, I have divided into various sub-sections. Lets jump right into it.


### Exploratory Data Analysis


The data comprises of fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

**mtcars is data frame with 32 observations on 11 variables**

Variable | Definition
------ | ----------------
mpg  |  Miles/(US) gallon  
cyl  |  Number of cylinders  
disp	|	Displacement (cu.in.)  
hp	|	Gross horsepower  
drat	|	Rear axle ratio  
wt	|	Weight (lb/1000)  
qsec	|	1/4 mile time  
vs	|	V/S  
am	|	Transmission (0 = automatic, 1 = manual)  
gear	|	Number of forward gears  
carb	|	Number of carburetors	


-------------------------------------------------------------------

We start by importing data into a new variable, `cars`, and then display first few observations of `cars`. 

```{r}
data(mtcars)
#importing data in a variable
cars <- mtcars
head(cars)
```

One may also run `summary(cars)` to get more interesting statistics about each variable. We observe that few variables such as 'Number of Cylinders', 'Number of forward gears', etc. are categorical variables. We convert those to [factors](http://www.stat.berkeley.edu/~s133/factors.html).


```{r}
# Changing classifying variables from numeric values to factor

cars$cyl <- factor(cars$cyl)
cars$vs <- factor(cars$vs)
cars$am <- factor(cars$am)
cars$gear <- factor(cars$gear)
cars$carb <- factor(cars$carb)
```



### Data Visualizations

Next, we visualize the data to understand the behaviour of the response variable. 


```{r, echo=FALSE}
par(mfrow = c(1, 2)) 
#visualize two graphs in one horizontal length
# Add a Normal Curve
x <- cars$mpg 
h<-hist(x, breaks=10, col="cornflowerblue", xlab="Miles Per Gallon", 
    main="Histogram with Normal Curve") 
xfit<-seq(min(x),max(x),length=40) 
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x)) 
yfit <- yfit*diff(h$mids[1:2])*length(x) 
lines(xfit, yfit, col="blue", lwd=2)

# Density Curve
plot(density(cars$mpg), xlab = "MPG", main ="Density Plot of MPG")
```

* Curve shows a near Normal Distribution


```{r, echo=FALSE, message=FALSE, warning=FALSE}
nums <- sapply(cars, is.numeric)

pairs(cars[,nums], main ="Finding Correlation among Variables", panel=function(x,y) {
  points(x, y)
  abline(lm(y ~ x), col="red")
})

```

* Above scatter plot matrix shows the correlation between each variables.
* We see a high correlation between response variables and predictor variables. 
* Also, we see  correlations among predictor variables. 


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plyr)
library(ggplot2)
# Rename the levels of transmission types
transmission <- revalue(cars$am, c('0'="automatic", '1'="manual"))
qplot(transmission, mpg, data=cars, geom=c("boxplot"), 
   fill=am, main="Mileage by Transmission Type",
   xlab="Transmission Type", ylab="Miles per Gallon")

```

From the box plot, we observe the following:

* Manual transmission is better than Automatic transmission.  
* We see the range of values and check for outliers. 

Code for the above plots are available on my github repository, you can fork the entire code [here](https://github.com/shubhamkalra27/mtcars-analysis/blob/master/mtcars.R). 

Next we will check if they are statistically significant using a [two sample t-test](http://www.itl.nist.gov/div898/handbook/eda/section3/eda353.htm). 
  
  
   
-----------------------------------  
  
### Two Sample t-test

Two-sample t-test is applied to compare whether the average difference between two groups is really significant or if it is due instead to random chance.

``` {r}
t.test(mpg~am, data = cars)
```

**Interpretation**   
From the above output, we reject the null hypothesis with a p-value of **0.001374**,  and say that there is a statistically significant difference in mean mileage between manual transmission and automatic transmission.

We will now Quantify that difference using Regression techniques.


###Simple linear regression

Our first regression model is to explain the `mpg` variability with the `am` only.  


``` {r}
model1 <- lm( mpg~am ,data=mtcars); summary(model1)
```


**Interpretation of simple linear regression**  

* Intercept value implies that cars with Automatic Transmission have on average  **`r round(summary(model1)$coefficients[1],3)`** MPG  
* Coefficient of the parameter `am` implies that Manual transmission cars add on average **`r round(summary(model1)$coefficients[2],3)`** MPG's more to the Automatic car's mean value  
* Adj R square value says, that only **`r round(summary(model1)$adj.r.squared*100,2)`%** of deviation in model can be explained by the current model with one predictor.
  


> Before making any conclusions on the effect of transmission type on mileage, we look at the impact of other variables on mileage



  
### Multiple Regression

Next, we would like to determine dependence of other variables on Mileage. To find out which preditor variables to use in our model, we run a Stepwise regression to identify useful subset of predictors. 

```{r, eval=FALSE }
nullModel <- lm (mpg ~ 1 , data =cars) # model with no predictor variables.

fullModel <- lm (mpg ~ . , data = cars) # model with all variables added as predictor variables. 

bestModel <- step(fullModel, nullModel, direction="both", k=log(nrow(cars))) #both direction BIC step model

bestModel$call  #display formula
```

```{r, echo=FALSE }
nullModel <- lm (mpg ~ 1 , data =cars) # model with no predictor variables.

fullModel <- lm (mpg ~ . , data = cars) # model with all variables added as predictor variables. 

bestModel <- step(fullModel, nullModel, direction="both", k=log(nrow(cars))) #both direction BIC step model

bestModel$call  #display formula
```


* The variables mentioned by bestModel are most useful set of predictors
* The final model `bestModel` contains three predictors, wt(weight), qsec(1/4 mile time) and am(Transmission)  
 

Summarizing the new Model.

```{r}
summary(bestModel)
```
  
  
  
**Interpretation of Multiple Regression**  

  
* Adjusted R square value signifies that **`r round(summary(bestModel)$adj.r.squared*100,2)`%** of the variance can be explained by this model, which is a significant improvement over first model.  
  
* 1000 pound increase in `wt`(weight) of a car, keeping all other variables constant, decreases MPG by **`r round(abs(summary(bestModel)$coefficients[2]),3)`**

* 1 unit change in `qsec`(1/4 mile time), with all other variables constant, increases MPG by **`r round(abs(summary(bestModel)$coefficients[3]),3)`**

* Coefficient for `am` signifies that, manual transmission cars have an average MPG of **`r round(summary(bestModel)$coefficients[4],3)`**  more than that of automatic transmission cars. 


* The p-value for `am` is `0.046` which makes it a weakest predictor of mileage.

-------------------------------------------------------------------

 
### Conclusion

Mileage (mpg), when modeled with just one predictor `am`, we get manual transmission to yield an average of `7.245`mpg higher than automatic transmission. Here, only **`r round(summary(model1)$adj.r.squared*100,2)`%** of variance could be explained.


Multiple regression on stepwise regressed variables defines **`r round(summary(bestModel)$adj.r.squared*100,2)`**% of variability. `wt` and `qsec` have [confounded](http://en.wikipedia.org/wiki/Confounding) the relationship between `am` and `mpg`. Manual transmission in this model increases mileage by only `2.936 mpg`  


-----------------------------------------  
  
#### References. 

1. mtcars dataset <https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html>
2. Density Plot. <http://www.statmethods.net/graphs/density.html>   
3. Analysis on mtcars dataset to predict mpg <http://varianceexplained.org/RData/code/code_lesson3/>
4. BoxPlot using ggplot2 <http://www.statmethods.net/advgraphs/ggplot2.html>
5. More on factors <http://www.stat.berkeley.edu/~s133/factors.html>  
6. Confounding Variables <http://en.wikipedia.org/wiki/Confounding/>
7. two-sample t test <http://www.isixsigma.com/tools-templates/hypothesis-testing/making-sense-two-sample-t-test/>
8. Coursera regmods <https://github.com/sefakilic/coursera-regmods/>

-------------------------------


You can access the code for the above analysis at my github repository. <https://github.com/shubhamkalra27/mtcars-analysis>

[follow author on Twitter](https://twitter.com/shubhamkalra27)
  

-------------------------------



